---
title: "Chicago Bears 2025 Offensive Analysis"
author: "Carson Kessler"
date: "`r Sys.Date()`"
output:
  html_document:
    theme:
      bootswatch: cosmo
      primary: "#0B162A"
    toc: true
    toc_float:
      collapsed: false
    toc_depth: 2
    number_sections: true
    code_folding: hide
    df_print: paged
    fig_width: 9
    fig_height: 5
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

```{css, echo=FALSE}
h1, h2, h3 {
  color: #0B162A;
  font-weight: 600;
}
body {
  font-size: 16px;
  line-height: 1.6;
}
.title {
  border-bottom: 4px solid #C83803;
  padding-bottom: 10px;
}
.summary-box {
  background-color: #f4f6f8;
  border-left: 5px solid #C83803;
  padding: 16px 20px;
  margin: 20px 0;
  border-radius: 4px;
}
```

## Introduction
This report analyzes the Chicago Bears offense during the 2025 NFL season using
play-by-play data from the nflreadr package. I look at EPA, success rate, and
situational performance to understand how the offense performed.

<div class="summary-box">
**Bottom line:** The 2025 Bears fielded a top-7 offense by EPA per play, built on
balanced, efficient early downs and strong third-down conversion. Their one clear
weakness was red zone finishing, where they ranked 24th of 32, which is the most
actionable area for improvement.
</div>

```{r load-packages}
library(nflreadr)
library(tidyverse)
library(plotly)
```

```{r theme-setup, include=FALSE}
theme_set(theme_minimal(base_size = 13))
theme_update(
  plot.title = element_text(face = "bold", color = "#0B162A"),
  plot.subtitle = element_text(color = "grey40")
)
```

```{r load-data}
pbp <- load_pbp(2025)
```

```{r clean-data}
offense <- pbp %>%
  filter(play_type %in% c("pass", "run"), !is.na(epa))
```
dim(offense)

## League-Wide Context: Where Did the Bears Rank?

Before zooming into the Bears specifically, it helps to see where they ranked
among all 32 teams in offensive EPA per play. This is one of the cleanest single
measures of overall offensive quality.

```{r team-epa-ranking}
team_epa <- offense %>%
  group_by(posteam) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa),
    .groups = "drop"
  ) %>%
  arrange(desc(epa_per_play)) %>%
  mutate(rank = row_number())

# Where did the Bears land?
team_epa %>% filter(posteam == "CHI")
```

The Bears finished 7th in the league in offensive EPA per play, placing them
firmly in the upper third of NFL offenses. The chart below shows the full league
ranking with Chicago highlighted.

```{r team-epa-plot}
p1 <- team_epa %>%
  ggplot(aes(x = reorder(posteam, epa_per_play), y = epa_per_play, fill = posteam == "CHI")) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "grey70", "TRUE" = "#0B162A")) +
  labs(
    title = "2025 Offensive EPA per Play by Team",
    subtitle = "Chicago highlighted in navy",
    x = NULL,
    y = "EPA per Play"
  ) + 
  theme_minimal() +
  theme(legend.position = "none")
ggplotly(p1, tooltip = "text")
```

## Pass vs Run Efficiency

A core question for any offense is how efficient they are throwing versus running.
Modern analytics generally finds passing more efficient than running, but the size
of that gap varies by team. Here's how the Bears split out.

```{r pass-run-split}
bears_split <- offense %>%
  filter(posteam == "CHI") %>%
  group_by(play_type) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa),
    success_rate = mean(success, na.rm = TRUE),
    avg_yards = mean(yards_gained, na.rm = TRUE),
    .groups = "drop"
  )
bears_split
```


The averages tell part of the story, but the distribution of outcomes reveals
more. The boxplot below shows the full spread of EPA for pass and run plays.

```{r pass-run plot}
p2 <- offense %>%
  filter(posteam == "CHI") %>%
  ggplot(aes(x = play_type, y = epa, fill = play_type)) + 
  geom_boxplot(alpha = 0.7) +
  scale_fill_manual(values= c("pass" = "#0B162A", "run" = "#C83803")) +
  labs(
    title = "Bears EPA Distribution: Pass vs Run",
    x = "Play Type",
    y = "EPA"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
ggplotly(p2)
```


The boxplot makes the tradeoff clear. Passing produced a much wider range of
outcomes, with explosive gains above +4 EPA but also costly negative plays below
-4. The run game clustered tightly around zero, offering consistency at the cost
of explosive upside. This explains why passing carried a higher average EPA
despite a lower success rate.

## Performance by Down

How an offense performs across downs reveals a lot about its identity. Strong
offenses stay efficient on early downs, which keeps them out of difficult
third-and-long situations and lets them stay unpredictable in their play calling.

```{r down-analysis}
bears_downs <- offense %>%
  filter(posteam == "CHI", down %in% 1:4) %>%
  group_by(down) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa),
    success_rate = mean(success, na.rm = TRUE),
    .groups = "drop"
    )
bears_downs
```
The Bears were highly efficient on first down (+0.135 EPA per play), which is the foundation of a good offense and a key reason they ranked 7th overall. Second down was their flattest down, hovering right at zero, suggesting they didn't consistently build on early-down success. They recovered on third down (+0.086 EPA, 46.6% success rate), showing an ability to convert and sustain drives. Fourth down was negative, but with only 41 plays the sample is too small to draw firm conclusions.

```{r down-plot}
p3 <- bears_downs %>%
  ggplot(aes(x = factor(down), y = epa_per_play, fill = epa_per_play > 0,
             text = paste0("Down: ", down,
                           "<br>EPA/play: ", round(epa_per_play, 3),
                           "<br>Plays: ", plays,
                           "<br>Success rate: ", round(success_rate * 100, 1), "%"))) +
  geom_col() +
  scale_fill_manual(values = c("FALSE" = "#C83803", "TRUE" = "#0B162A")) +
  labs(title = "Bears EPA per Play by Down",
       x = "Down", y = "EPA per Play") +
  theme_minimal() +
  theme(legend.position = "none")

ggplotly(p3, tooltip = "text")
```


## Play Calling Tendencies

This section looks at how pass-heavy or run-heavy the Bears were across downs,
which reveals their offensive philosophy and how predictable they may have been to
opposing defenses.

```{r tendencies}
bears_tendency <- offense %>%
  filter(posteam == "CHI", down %in% 1:3) %>%
  group_by(down) %>%
  summarize(
    pass_rate = mean(play_type == "pass") * 100,
    .groups = "drop"
  )
bears_tendency
```

The Bears were remarkably balanced on early downs, splitting exactly 50/50 between
pass and run on first down and staying near even on second down. This balance made
them difficult to predict early in the down sequence. As expected, they shifted
pass-heavy on third down (65.4%), when distance-to-go typically demands it. The
early-down balance, combined with their strong first-down efficiency, suggests an
offense that stayed ahead of schedule and kept defenses honest rather than tipping
its hand.

```{r tendency-plot}
p4 <- bears_tendency %>%
  ggplot(aes(x = factor(down), y = pass_rate,
             text = paste0("Down: ", down,
                           "<br>Pass rate: ", round(pass_rate, 1), "%"))) +
  geom_col(fill = "#0B162A") +
  geom_text(aes(label = paste0(round(pass_rate, 1), "%")),
            nudge_y = 4, color = "black") +
  labs(title = "Bears Pass Rate by Down",
       x = "Down", y = "Pass Rate (%)") +
  ylim(0, 100) +
  theme_minimal()

ggplotly(p4, tooltip = "text")
```


## Situational Analysis: Third Down and Red Zone

Third down conversion and red zone efficiency are two of the most important
situational metrics in football. They directly translate to sustaining drives and finishing them with points, and they often separate good offenses from great ones.

```{r third_down}
third_down <- offense %>%
  filter(down == 3, !is.na(success)) %>%
  group_by(posteam) %>%
  summarize(conversion_rate = mean(success) * 100, .groups = "drop")

bears_third <- third_down %>% filter(posteam == "CHI") %>%
  pull(conversion_rate)
league_third <- mean(third_down$conversion_rate)

cat("Bears 3rd down success rate:", round(bears_third, 1), "%/n")
cat("League average:", round(league_third, 1), "%/n")
```
The Bears converted on third down at a 46.6% clip, comfortably above the leagu average of 42.3%. Third down is where drives are extended or killed, so finishing above average here reflects an offense that sustained possessions well. This lines up with their third-down EPA recovery seen earlier.

```{r red-zone}
red_zone <- offense %>%
  filter(yardline_100 <= 20) %>%
  group_by(posteam) %>%
  summarize(
    rz_plays = n(),
    tds = sum(touchdown == 1, na.rm = TRUE),
    td_rate = tds / rz_plays * 100,
    .groups = "drop"
  )
bears_rz <- red_zone %>% filter(posteam == "CHI")
bears_rz
```


```{r red-zone-context}
red_zone <- red_zone %>%
  arrange(desc(td_rate)) %>%
  mutate(rank = row_number())

league_rz <- mean(red_zone$td_rate)

red_zone %>% filter(posteam == "CHI")
cat("League average red zone TD rate (per play):", round(league_rz, 1), "%/n")
```
Red zone finishing was the clear weak spot. The Bears ranked 24th of 32 in red
zone touchdown rate per play (17.3%), below the league average of 19.1%. This
stands out against their otherwise strong profile: a top-7 offense that moved the
ball efficiently between the 20s but struggled to finish drives with touchdowns,
likely settling for field goals where better red zone offenses found the end zone. This is the most actionable takeaway in the analysis and points to a concrete area for improvement.
  
## EPA Over the Course of the Season

Tracking EPA per play by week shows whether the offense improved, declined, or
stayed steady as the season progressed. This can reveal the impact of coaching
adjustments, player development, injuries, or strength of schedule.

```{r weekly-trend}
weekly_epa <- offense %>%
  filter(posteam == "CHI") %>%
  group_by(week) %>%
  summarize(epa_per_play = mean(epa), .groups = "drop")

p5 <- weekly_epa %>%
  ggplot(aes(x = week, y = epa_per_play,
             text = paste0("Week: ", week,
                           "<br>EPA/play: ", round(epa_per_play, 3)))) +
  geom_line(color = "#0B162A", linewidth = 1, group = 1) +
  geom_point(color = "#C83803", size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey50") +
  labs(title = "Bears Weekly Offensive EPA per Play",
       x = "Week", y = "EPA per Play") +
  theme_minimal()

ggplotly(p5, tooltip = "text")
```

The weekly trend reveals a volatile offense. The Bears swung between elite
performances (peaking around +0.45 EPA per play in Week 9, with other big games in
Weeks 3 and 17) and quieter weeks that dipped below league-neutral. Most weeks
landed above the zero line, consistent with their strong full-season ranking, but
the large week-to-week swings show an offense that was more boom-or-bust on a game
level than steadily dominant. This echoes the boom-or-bust tendency seen in their
passing distribution.


```{r offensive-points}
bears_points <- pbp %>%
  filter(posteam == "CHI") %>%
  group_by(week) %>%
  summarize(
    off_tds = sum(touchdown == 1 & td_team == "CHI", na.rm = TRUE),
    field_goals = sum(field_goal_result == "made", na.rm = TRUE),
    extra_points = sum(extra_point_result == "good", na.rm = TRUE),
    two_pt = sum(two_point_conv_result == "success", na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    offensive_points = off_tds * 6 + field_goals * 3 + extra_points * 1 + two_pt * 2)
bears_points %>% select(week, off_tds, field_goals, offensive_points)
```

```{r offensive-points-plot}
p6 <- bears_points %>%
  ggplot(aes(x = week, y = offensive_points, 
             text = paste0("Week: ", week,
                           "<br>Offensive points: ", offensive_points,
                           "<br>Offensive TDs: ", off_tds,
                           "<br>Field Goals: ", field_goals))) +
  geom_col(fill = "#0B162A") +
  geom_hline(yintercept = mean(bears_points$offensive_points),
             linetype = "dashed", color = "#C83803") +
  labs(
    title = "Bears Offensive Points Scored by Week",
    subtitle = "Dashed line marks season average (offensive points only)",
    x = "Week",
    y = "Offensive Points"
  ) +
  theme_minimal()
ggplotly(p6, tooltip = "text")
```
## Offensive Scoring Output

Efficiency metrics like EPA capture how well an offense moves the ball, but points
are what ultimately decide games. This section isolates offensive points only,
excluding any scores from defense or special teams, to measure how the unit's
efficiency translated into actual production.

```{r quarter-analysis}
bears_quarters <- offense %>%
  filter(posteam == "CHI", qtr %in% 1:4) %>%
  group_by(qtr) %>%
  summarize(
    plays = n(),
    epa_per_play = mean(epa, na.rm = TRUE),
    success_rate = mean(success, na.rm = TRUE),
    pass_rate = mean(play_type == "pass") * 100,
    .groups = "drop"
  )
bears_quarters
```

```{r quarter-plot}
p7 <- bears_quarters %>%
  ggplot(aes(x = factor(qtr), y = epa_per_play, fill = epa_per_play > 0,
                        text = paste0("Quarter: ", qtr,
                                      "<br>EPA/play: ", round(epa_per_play, 3),
                                      "<br>Plays: ", plays,
                                      "<br>Success rate: ", round(success_rate * 100, 1), "%",
                                      "<br>Pass rate: ", round(pass_rate, 1), "%"))) +
           geom_col() +
           scale_fill_manual(values = c("FALSE" = "#C83803", "TRUE" = "#0B162A")) +
           labs(title = "Bears EPA per Play by Quarter",
                x= "Quarter", y = "EPA per Play") +
           theme_minimal() +
           theme(legend.position = "none")
         ggplotly(p7, tooltip = "text")
```

The quarter splits reveal a story the season averages completely hide: the Bears
were a second-half team. They were neutral in the first quarter (+0.019 EPA) and
actually negative in the second (-0.013), their only negative quarter, before
surging after halftime to +0.101 in the third and a season-best +0.168 in the
fourth. The jump from the second to third quarter suggests effective halftime
adjustments, and the strong fourth quarter, paired with a higher pass rate (58.8%),
points to an offense that was dangerous late in games. Some of the fourth-quarter
production likely reflects favorable game script when trailing, but the strong
third quarter, when game state is typically more neutral, suggests this was a
genuine pattern rather than a garbage-time artifact. The clear takeaway for the
coaching staff: if the Bears could start games the way they finished them, this
offense has another level. The slow starts were arguably their biggest missed
opportunity.

## Methodology

Data comes from the nflverse `nflreadr` package, covering all 2025 regular season
plays. The analysis is limited to run and pass plays with valid EPA values,
excluding special teams and plays with missing data. EPA (Expected Points Added)
measures how much each play changed the team's expected points given the game
situation, and success rate captures the share of plays that gained enough yardage
to be considered successful for the down and distance.

## Key Takeaways

The 2025 Chicago Bears offense was strong but uneven. The headline findings:

- **Top-tier overall efficiency.** The Bears ranked 7th of 32 in offensive EPA per
  play (+0.074), placing them firmly in the upper third of the league.

- **Balanced and unpredictable early downs.** A perfect 50/50 run-pass split on
  first down, paired with strong first-down EPA (+0.135), kept defenses honest and
  the offense ahead of schedule.

- **Passing drove the upside.** Passing carried higher EPA than running despite a
  lower success rate, reflecting a boom-or-bust passing attack that produced
  explosive plays but also negative ones.

- **Above-average on third down.** A 46.6% third-down success rate, well above the
  42.3% league average, showed an ability to sustain drives.

- **Red zone finishing was the weak spot.** Ranking 24th of 32 in red zone
  touchdown rate, the Bears struggled to convert scoring chances into touchdowns,
  the clearest area for improvement.

- **Volatile week to week.** The offense alternated between elite and quiet
  performances rather than producing steady output.

**Overall:** This was a quality offense, top-7 by the numbers, whose ceiling was
limited by red zone finishing and game-to-game inconsistency. The most actionable
takeaway for the coaching staff would be improving red zone efficiency, where even
league-average finishing would have meaningfully increased scoring output.